---
title: "432_group1_A2"
author: "Emma Mitchell"
date: "2025-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Emma Mitchell - 20296602 

Load the Needed Packages:

```{r}
library(dplyr)
```

```{r}
library(tidyverse)
```

```{r}
library(ggplot2)
```

```{r}
library(tidyr)
```

```{r}
data<-read.csv("/Users/emmamitchell/Downloads/BirdBehaviour.csv")
```
Check data integrity: 

```{r}
class(data)
dim(data)
summary(data)
sum(is.na(data))
head(data)
tail(data)
```

no na's 

Create a multi-panel figure to investigate how the four behavioural measurements differ among bird species. You must choose what kind of plots to use here. You may want to try a few different options and then choose the one that best communicates the results. Be sure to include a figure caption.


```{r}
ggplot(data, aes(x = Species, y = ))
```



Calculate the correlation coefficient between each pair of behaviours. 



Add a paragraph of text explaining your interpretation. What do these figures and correlations suggest to you about the behaviour of these birds?









